\documentclass[conference]{IEEEtran}

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{cite}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{array}
\usepackage{xurl}
\usepackage[hidelinks]{hyperref}
\Urlmuskip=0mu plus 1mu

\title{Hardware Security}

\author{\IEEEauthorblockN{Author}
\IEEEauthorblockA{Hardware Security Survey}}

\begin{document}

\maketitle

\begin{abstract}
This survey reviews defense mechanisms for Arm and RISC-V confidential
computing, confidential-computing network and I/O data paths, and ISA-level
hardware design defenses. The central observation is that modern hardware
security cannot be evaluated by CPU isolation alone. A confidential workload is
protected only when execution domains, physical-memory ownership, device
assignment, DMA translation, interrupt routing, link protection, and attestation
evidence compose into a coherent platform boundary. The survey therefore
connects Arm CCA, RISC-V CoVE/AP-TEE, trusted I/O protocols, accelerator TEEs,
and hardware mechanisms such as MTE, PMP, capability machines, and CFI
extensions under a common systems-oriented taxonomy.
\end{abstract}

\begin{IEEEkeywords}
hardware security, confidential computing, trusted execution environments,
Arm CCA, RISC-V, attestation, I/O protection
\end{IEEEkeywords}

\section{Introduction}

Hardware security has shifted from isolated processor features toward
platform-level protection contracts. Modern systems must coordinate execution
domains, memory ownership, device access, interrupt delivery, and attestation
evidence across the whole machine. This survey organizes the current literature
around that composition problem, using Arm and RISC-V as the main architectural
lines.

The scope is deliberately narrower than a general hardware-security SoK. We
focus on three defense-oriented lanes. The first lane covers confidential
computing mechanisms in Arm and RISC-V systems, including Arm TrustZone and CCA
as well as RISC-V enclave lineage, CoVE, AP-TEE, and TVM lifecycle management.
The second lane covers network, I/O, and data-path defense in
confidential-computing deployments, including SPDM, TDISP, PCIe IDE,
IOMMU/IOPMP/SMMU mediation, trusted MSI, SmartNIC/DPU-style offload, and
accelerator TEEs. The third lane covers ISA and hardware design defenses that
provide defense-in-depth for those platforms: memory tagging, page-table
permissions, PMP/ePMP, capability/tagging architectures, hardware CFI,
memory-encryption taxonomy, and debug or trace controls.

This scope also defines what the survey does not attempt to cover in the
current version. Side channels, physical leakage, fault injection, Rowhammer,
and attack-only vulnerability catalogs are important hardware-security topics,
but they are not treated as independent research lanes here. They are mentioned
only when they affect the threat-model boundary of a confidential-computing
mechanism. Similarly, ordinary network-security mechanisms such as firewalls,
generic IDS/IPS systems, routing security, TLS, or VPNs are excluded unless they
directly protect a confidential workload's trusted network endpoint, device
assignment path, or secure offload boundary.

The paper is organized as follows. The Arm section explains the transition from
TrustZone-style TEEs to CCA and Realm-based protection. The RISC-V section
connects earlier enclave systems to CoVE/AP-TEE confidential VMs. The
confidential I/O section then treats devices, fabrics, and network data paths
as first-class parts of the confidential-computing boundary. Finally, the
hardware-design section surveys cross-architecture ISA and microarchitectural
defenses that complement those confidential-computing mechanisms.

\section{Methodology and Evidence Ranking}

This survey separates evidence strength from topical relevance. A source may
be important to the taxonomy while still being unsuitable for a mechanism-level
claim. We therefore use the following evidence classes when deciding how a
reference may support the text:

\begin{table}[htbp]
\centering
\caption{Evidence Classes and Citation Rules}
\label{tab:evidence_ranking}
\begin{tabular}{@{}p{0.12\linewidth}p{0.36\linewidth}p{0.42\linewidth}@{}}
\toprule
\textbf{Class} & \textbf{Evidence type} & \textbf{Allowed use in this survey} \\ \midrule
E0 & Official ratified specification, architecture manual, RFC, or standards document & Normative mechanism and status claims, with no experimental claims unless the source provides them. \\
E1 & Peer-reviewed primary research paper & Mechanism, evaluation, and threat-model claims supported by the reviewed paper and its recorded source/PDF status. \\
E2 & Survey, SoK, or secondary review & Taxonomy, terminology, and coverage guidance; original mechanism claims are traced back to primary papers or specifications. \\
E3 & Public draft, release candidate, or not-ratified specification & Current standardization direction, always labelled as draft, release candidate, or not ratified. \\
E4 & Vendor, industry, or product documentation & Industry evidence for product behavior or deployment practice, not proof of general security unless independently supported. \\
E5 & Metadata-only, gated, blocked-PDF, or HTML-only evidence & Existence, source status, or backlog classification only; no substantive mechanism claim unless promoted or paired with stronger public evidence. \\ \bottomrule
\end{tabular}
\end{table}

The citation rules are intentionally conservative. First, all mechanism claims
prefer E0 or E1 sources. Second, E2 sources such as surveys and SoKs are used as
maps of the literature, not as substitutes for original mechanisms. Third, E3
drafts are useful for current architecture direction, but the text must state
their draft or not-ratified status. Fourth, E4 industry documents are treated
as deployment or product evidence and kept separate from peer-reviewed system
claims. Fifth, E5 sources, including metadata-only \texttt{rw...} candidates and
member-gated documents, do not support substantive text claims unless a public
source supports the same claim. When a source has mixed properties, such as a
public metadata page for a gated specification or an HTML-only survey, the more
restrictive citation rule controls the claim.

\begin{table*}[t]
\centering
\scriptsize
\caption{Key Platform and Standard Sources}
\label{tab:key_platform_standard_sources}
\begin{tabular}{@{}p{0.24\linewidth}p{0.09\linewidth}p{0.24\linewidth}p{0.33\linewidth}@{}}
\toprule
\textbf{Key} & \textbf{Class} & \textbf{Status/version} & \textbf{Allowed claim in this survey} \\ \midrule
\texttt{arm\_cca\_spec} & E0 & Arm official specification, 2025 & CCA architecture semantics, Realm security state, and architectural protection model. \\
\texttt{arm\_rme\_spec} & E0 & Arm official RME specification, 2025 & RME, GPT/GPC, granule ownership, and physical-address-space semantics. \\
\texttt{arm\_rmm\_spec} & E0 & Arm official RMM specification, 2025 & RMM lifecycle, RMI/RSI interface roles, and monitor-side CCA state. \\
\texttt{li2022cca} & E1 & OSDI 2022 primary paper & Arm CCA design and verification claims supported by the paper. \\
\texttt{arm\_smmu\_spec} & E0 & Arm SMMUv3 architecture specification & DMA translation and system I/O mediation substrate, not complete trusted I/O by itself. \\
\texttt{riscv\_privileged} & E0 & RISC-V privileged architecture & Privilege modes, PMP/ePMP/Smepmp, virtualization substrate, and page-table semantics. \\
\texttt{riscv\_iommu\_2023} & E0 & RISC-V IOMMU v1.0.1 / 20260222 ratified-library release & Device-side DMA translation substrate for confidential-I/O composition. \\
\texttt{riscv\_aia\_2023} & E0 & RISC-V AIA v1.0, ratified June 2023; revised 20250312 clarifications & AIA/IMSIC interrupt substrate and trusted-MSI vocabulary. \\
\texttt{riscv\_iopmp\_2026} & E3 & IOPMP v0.8.2 draft, not ratified & Current I/O protection direction; draft status must remain visible. \\
\texttt{sahita2023cove} & E3 & CoVE public preprint / proposal & CoVE motivation and reference-architecture direction, not ratified ABI detail. \\
\texttt{riscv\_ap\_tee\_2024} & E3 & Draft/not ratified: AP-TEE v0.7 / RC2 & RISC-V TVM/TSM lifecycle and ABI direction, labelled as draft. \\
\texttt{riscv\_cove\_io\_2026} & E3 & Draft/not ratified: CoVE-IO v0.3.0 & Trusted-I/O architecture direction for RISC-V TVMs, labelled as draft. \\
\texttt{dmtf\_spdm\_2025} & E0 & DMTF DSP0274 v1.4.0, selected as this survey's SPDM snapshot & SPDM identity, measurement, negotiation, and key-exchange semantics. \\
\texttt{pcisig\_tdisp\_2022} & E5 & Public ECN metadata; member-gated PDF & TDISP existence, affected-specification metadata, and high-level device-interface role only. \\
\texttt{pcisig\_tdisp\_xt\_2025} & E5 & Public PCIe 6.x ECN metadata; member-gated download & TDISP XT public existence and affected IDE/ATS/TDISP metadata only; no gated XT/TE normative-rule claims. \\
\texttt{eat\_rfc} & E0 & RFC 9711, Standards Track & EAT claim/token encoding for attestation evidence. \\
\texttt{rats\_rfc} & E0 & RFC 9334, Standards Track & RATS attester, verifier, relying-party, evidence, and appraisal roles. \\ \bottomrule
\end{tabular}
\end{table*}

\begin{table*}[t]
\centering
\scriptsize
\caption{Key Survey, Vendor, and Boundary Sources}
\label{tab:key_boundary_sources}
\begin{tabular}{@{}p{0.24\linewidth}p{0.09\linewidth}p{0.24\linewidth}p{0.33\linewidth}@{}}
\toprule
\textbf{Key} & \textbf{Class} & \textbf{Status/version} & \textbf{Allowed claim in this survey} \\ \midrule
\begin{tabular}[t]{@{}l@{}}\texttt{nvidia\_bluefield}\\\texttt{operation\_2025}\end{tabular} & E4 & NVIDIA BlueField public documentation & DPU OP-TEE/fTPM building-block evidence only, not complete confidential networking proof. \\
\texttt{boubakri2025riscvtee} & E2/E5 & Survey source; local PDF blocked / source verified & RISC-V TEE taxonomy support only; mechanism claims return to primary papers or specs. \\
\texttt{schneider2022soktee} & E2 & Hardware-supported TEE SoK / survey & Cross-platform TEE design-space taxonomy and terminology. \\
\texttt{li2024sokteechoices} & E2 & ASIA CCS 2024 SoK & TEE runtime design-choice and pitfall taxonomy, not platform-specific mechanism proof. \\
\texttt{sok-tee} & E2 & Accelerator TEE SoK & Accelerator/device-TEE taxonomy; not general CCA or CoVE mechanism evidence. \\
\texttt{perkins2024socsok} & E2 & SoC-FPGA TEE SoK & SoC-FPGA accelerator-TEE coverage and gap map; primary mechanism claims return to original systems. \\
\texttt{fan2025xputee} & E5 & ACM TOCS metadata; local PDF unavailable & Recent CPU/GPU heterogeneous-TEE baseline at abstract level only; no strong mechanism or performance claim. \\
\texttt{xia2021sgxfpga} & E5 & DAC metadata; local PDF unavailable & CPU-FPGA TEE lineage marker only; no strong mechanism claim without the paper PDF. \\
\texttt{feng2024siopmp} & E1 & ASPLOS 2024 primary paper & sIOPMP mechanism and evaluation claims for scalable I/O protection. \\ \bottomrule
\end{tabular}
\end{table*}

Tables~\ref{tab:key_platform_standard_sources} and~\ref{tab:key_boundary_sources}
make the strongest sources and the main boundary cases visible in the paper
body. In particular, draft RISC-V specifications, vendor documents, surveys,
and gated PCI-SIG metadata are listed with their restricted claim roles rather
than being treated as ordinary mechanism evidence.

\input{tee_design_space}

\input{confidential_computing_of_arm_hardware}

\input{riscv_confidential_computing}

\input{confidential_io_and_network_defense}

\input{security_of_hardware_design}

\bibliographystyle{IEEEtran}
\bibliography{reference}

\end{document}
